### 2020-07-11 paper.tex
checkpoint on probabilistic programming with traces

\section{INVOLUTIVE MCMC}
Let $p$ denote the target probability density with respect to some $\sigma$-finite measure $\mu_X$ on a set $X$.
For each $x$ such that $p(x) > 0$, let $q_x$ denote a probability density with respect to some $\sigma$-finite measure $\mu_U$ on a set $U$.
Let $\mu := \mu_X \times \mu_U$ denote the product measure on the space $X \times U$, and let $\pi(x, u) := p(x) q_x(u)$ and $Z := \{(x, u) : \pi(x, u) > 0\}$.
Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.

@@

%\begin{lemma}\label{lemma:involution-detailed-balance} % TODO cite
%For $p$, $q$, and $f$ satisfying the properties above, let $k'$ denote the following transition kernel on $Z$:
%\[
%k'_z(A) := [f(z) \in A] \alpha(z) + [z \in A](1 - \alpha(z))
%\]
%The transition kernel $k'$ on $Z$ defined by the involution $f$ satisfies detailed balance with respect to $\pi$:
%\[
%\int_B k'_z(A) \pi(dz) = \int_A k'_z(B) \pi(dz) \mbox{ for all } A, B \in \Sigma_Z
%\]
%And in particular it is stationary with respect to $\pi$:
%\[
%\int_Z k'_z(A) \pi(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
%\]
%\end{lemma}
%\begin{proof}
%\end{proof}
%
%\begin{theorem}\label{theorem:stationarity}
%The transition kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
%\[
%k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
%\]
%is stationary with respect to $p$.
%\end{theorem}
%\begin{proof}
%By Lemma~\citep{lemma:involution-detailed-balance}, the transition kernel $k'$ is stationary with respect to $\pi$:
%\[
%%\int_B k'_{z}(B) \pi(z) \mu(dz) = \pi(B) \mbox{ for all } z \in X \times U, B \in \Sigma_Z
%\]
%It suffices to show that
%\begin{align*}
%\int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
%\int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
%\int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)
%\end{align*}
%By Lemma~\citep{lemma:involution-detailed-balance}, setting $B := X \times U$, the left-hand-side is:
%\[
%\pi(X \times U) = \int_X \int_U p(x) q_x(u) \mu(
%\]
%\end{proof}
%\paragraph{Detailed balance of the involution}
%%Let $q_
%
%\paragraph{Stationarity}
%
%\paragraph{Conditional distribution and observations fixed}

@@

\subsection{Exploiting Jacobian Sparsity}

@@

\subsection{Proof of detailed balance for involution} \label{sec:involution-detailed-balance}
\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
We reproduce the result in our notation:
\begin{lemma}[Detailed balance for involution move~\citep{tierney1998note}]
Let $(Z, \Sigma, \pi)$ denote a measure space.
Suppose $f$ is a one-to-one function from $Z$ onto $Z$ such that $f^{-1} = f$.
Consider the probability kernel $k$ defined by $k_z(A) := [f(z) \in A] \alpha(z, f(z)) + [z \in A] (1 - \alpha(z, f(z)))$ (where $\alpha(z, f(z))$ gives the probability of accepting a proposed transition from $z$ to $f(z)$).
Let $\nu(dz) := \pi(dz) + (\pi \circ f^{-1})(dz)$.
Let $h(z)$ be a density for $\pi$ with respect to $\nu$.
Let $A := \{z \in Z : h(z) > 0 \mbox{ and } h(f(z)) > 0\}$.
$k$ satisfies detailed balance with respect to $\pi$ if and only if:
\begin{enumerate}
\item $\alpha(z, f(z)) = 0$ for $\pi$-almost all $z \not \in A$
\item $\alpha(z, f(z)) \frac{h(z)}{h(f(z))} = \alpha(f(z), z)$
\end{enumerate}
\end{lemma}

@@

>>>>

\section{INVOLUTIVE MCMC}

% problem: what if the return value of p determines the control flow of q_x?
% then the reference measure on 
%Let $(Z,

Let $p$ denote the target probability density with respect to some $\sigma$-finite measure $\mu_X$ on a set $X$.
For each $x$ such that $p(x) > 0$, let $q_x$ denote a probability density with respect to some $\sigma$-finite measure $\mu_U$ on a set $U$.
Let $\mu := \mu_X \times \mu_U$ denote the product measure on the space $X \times U$, and let $\pi(x, u) := p(x) q_x(u)$ and $Z := \{(x, u) : \pi(x, u) > 0\}$.
Let $f : Z \to Z$ denote an involution ($f^{-1} = f$) such that the pushforward of $\mu$ under $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$, with Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d\mu : Z \to [0, \infty)$.

@@

\section{PROBABILISTIC PROGRAMMING WITH TRACES}

Probabilistic programs are programs that define probability distributions.
Probabilistic programs can be used to define standard joint probability distributions over vectors of random-variables, as well as more complex probability distributions with uncertainty over the structure in which the set of random variables itself random.
Involutive MCMC can be applied for target probability distributions on non-standard reference measures, including models with structure uncertainty.
For example, reversible jump MCMC, which is a special case of involutive MCMC, can be applied to switch between different model structures.
This section describes a framework for probabilistic programming that is expressive enough to represent probability distributions with structure uncertainty.

\paragraph{Traces}
In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the name of a `random choice' to its value.
Following standard terminology in probabilistic programming~\citep{?}, we call these dictionaries \emph{traces} because they can be sampled by running a probabilistic program, and `tracing' the execution the program by recording the value of eacn random choice encountered.
In the probabilistic programming language used in this paper, the user assigns a unique name or \emph{address} to each random choice using a special syntax:

TODO: include example of tilde syntax

TODO: introduce the mathematical notation for a trace

\paragraph{A probabilistic program encodes a probability distribution on traces}
For probabilistic programs that only sample individual random choices from discrete probability distributions, the meaning of the program can be described by a probability mass function that assigns a probailtiy to every possible trace.

TODO: Give a simple example of a probabilistic program with discrete random chocies and stochastic control flow, and the resulting probabiilty distribution.
TODO: describe the two variants of tilde syntax

Note that traces are always finite, but there can be no upper bound.
For example, for this program, there is a countably infinite number of possible traces:

TODO: show a discrete probabilistic prgram with the table of traces it could sample

% measure version
\paragraph{Handling continuous distributions}
For programs that can sample from continuous probability distributions, the formalism needs to be extended.
pecifically, we define the $\mathcal{S}$ to be the set of all possible \emph{trace structures} as a (possibly countably infinite) set of finite sets of addresses, denoted $\mathcal{S} \subseteq P(\mathcal{A})$, where $\mathcal{A}$ is the set of all addresses.
For each address $a \in \mathcal{A}$ we assign a reference measure space $(X_a, \Sigma_a, \mu_a)$ where $\mu_a$ is $\sigma$-finite.
The set of all traces of a probabilistic program is the set $\mathcal{T} := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a\}\}$.
The \emph{reference measure space} for a probabilistic program is $(\mathcal{T}, \Sigma, \mu)$ where
$\Sigma$ is the $\sigma$-algebra that contains sets $\cup_{S \in \mathcal{S}}\{ B_S \} \in \mathcal{P}(\mathcal{T})$ such that $B_S$ is an element of the product $\sigma$-algebra $\Sigma_S := \otimes_{a \in S} \Sigma_a$,
and where $\mu(\cup_{S \in \mathcal{S}}\{ B_S \}) := \sum_{S \in \mathcal{S}} \mu_S(B_S)$ where $\mu_S$ is the product measure $\times_{a \in S} \mu_a$.

We assume that a probabilistic program denotes a reference measure space $\mu$ and a density $p(t)$ on traces $t$ with respect to the reference measure, such that $p$ and $\mu$ are mutually absolutely continuous.
We do not give the denotational semantics of a probabilistic programming language with this property, but instead refer to POPL paper for an example of a denotational semantics. % TODO what about requirement that the support of each random choice is constant? (this is implied by the definition of reference measure)

TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and the vbase measure and a density for it.

% step 1: *assert* that the composition of p and then q defines a probabilistic program with ref meas mu and mut. abs. cont. pi(z) [ok] (( do this without ascribing separate semantics to p and q?? ))
% step 2: show detailed balance on the joint space ... [ok]
% step 3: show stationarity on the joint space [ok, easy from detailed balance]
% step 4: show stationarity with respect to space of p.... ((requires ascribing semantics to p, but maybe not q??)
% - this would be straightforward for a product space, but our space is more complicated?
% - need to assert relationship between p and pi, ...

\[
p(t) \;\; \pi(z)
\]

\paragraph{Extending one probabilistic program with another}
Given a

% the only place where this matters is in the 'gibbs move' argument.. (8.4)

We need $p$ to be a density with respect to some reference measure space $(T_P, \Sigma_{P}, \mu_P)$.
We need $q_t$ to be a density with respect to some reference measure $\mu_{Q,t}$ for each $t \in T_P$.
Such that the function $\pi(t, u) := p(x) q_t(u)$ is a density $\pi(x, u)$ with respect to some reference measure $\mu$, and $\pi$ is mutually absolutely continuous with $\mu$.
(this can be constructed by using hierarchical addresses `model' and `proposal' for the random choics made in the two programs).

\paragraph{Conditioning}

\section{DIFFERENTIABLE PROGRAMMING WITH TRACES}

@@

@@

\subsection{Proof of detailed balance for involution} \label{sec:involution-detailed-balance}
\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
We reproduce the result in our notation:
\begin{lemma}[Detailed balance for involution move~\citep{tierney1998note}] \label{lemma:tierney-involution}
Let $(Z, \Sigma, \pi)$ denote a measure space.
Suppose $f$ is a one-to-one function from $Z$ onto $Z$ such that $f^{-1} = f$.
Consider the probability kernel $k$ defined by $k_z(A) := [f(z) \in A] \alpha(z, f(z)) + [z \in A] (1 - \alpha(z, f(z)))$ (where $\alpha(z, f(z))$ gives the probability of accepting a proposed transition from $z$ to $f(z)$).
Let $\nu(dz) := \pi(dz) + (\pi \circ f^{-1})(dz)$.
Let $h(z)$ be a density for $\pi$ with respect to $\nu$.
Let $A := \{z \in Z : h(z) > 0 \mbox{ and } h(f(z)) > 0\}$.
$k$ satisfies detailed balance with respect to $\pi$ if and only if:
\begin{enumerate}
\item $\alpha(z, f(z)) = 0$ for $\pi$-almost all $z \not \in A$
\item $\alpha(z, f(z)) \frac{h(z)}{h(f(z))} = \alpha(f(z), z)$
\end{enumerate}
\end{lemma}

@@

\paragraph{Proof of stationarity}
We need the composition of the two programs to define a valid probabilistic program with density $\pi$ that is absolutely continuous with respect to the refernce measure $\mu$.
We also need there to be a particular relationship between $\pi$ and $p$.
The involution part of the kernel is a kernel on the combined space $(Z, \Sigma_Z, \mu)$.
\[
k'_z(B_Z) = [f(z) \in B_Z] \alpha(z) + [z \in B_Z] (1 - \alpha(z))
\]
The kernel of Algorithm 1 is defined by:
\[
k_t(B_T) := \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du)
\]
We want to show it is stationary:
\begin{align}
\int_T k_t(B_T) p(t) \mu_P(dt) &= p(B_T) \mbox{ for all } B_T \in \Sigma_P\\
\int_T \left( \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du) \right) p(t) \mu_P(dt) &= p(B_T)\\
\sum_{i} \int_{T_C} \left( \int_{U_t} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,t}(du) \right) p(i, t_c) \mu_{P_C,i}(dt_c) &= p(B_T)\\ % separate out discrete part of t (i) and continuous part of t (t_c)
\sum_{i} \int_{T_C} \left( \int_{U_i} q_{t}(u) k'_{(t, u)}(\{(t', u') \in Z : t' \in B_T\}) \mu_{Q,i}(du) \right) p(i, t_c) \mu_{P_C,i}(dt_c) &= p(B_T)\\ % use that the reference measure mu_{Q,t} only depends on the discrete part
..\\
\int_{Z} \pi(t, u) k'_{t,u}(\{(t', u') \in Z : t' \in B_T\}) \mu(dz) &= p(B_T)\\
\pi(\{(t', u') \in Z : t' \in B_T\}) &= p(B_T)\\
\end{align}

the set of addresses $\mathcal{A}$ is partitioned into two spaces $\mathcal{A}_1$ and $\mathcal{A}_2$\\

each structure $S$ partitions into two structures $S_1$ and $S_2$

then $Z := \cup_{S \in \mathcal{S}} \{(S, \mathbf{x}) : \mathbf{x} \in \times_{a \in S} X_a \}$\\

Space for $Z$:
\[
\cup_{S \in \mathcal{S}} \{((\mathcal{A}_1 \cap S) \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in \times_{a \in S_1} X_a, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
\]

Space for $T$:
\[
\cup_{S_1 \in \mathcal{S}} \{(S_1, \mathbf{x_1}) : \mathbf{x_1} \in \times_{a \in S_1}\}
\]

\begin{align}
\pi(
\cup_{S \in \mathcal{S} : A_1 \cap S = S_1} \{(S_1 \cup (\mathcal{A}_2 \cap S), \mathbf{x_1}, \mathbf{x_2}) : \mathbf{x_1} \in B_{S_1}, \mathbf{x_2} \in \times_{a \in S_2} X_a\}
) &= p(B_T)\\
\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} \pi(S_1, S_2, \mathbf{x}_1, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1} \times \otimes_{a \in S_2} X_a} p(S_1, \mathbf{x}_1) q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_1 \cup S_2}(d \mathbf{x}_1 \mathbf{x}_2) &= p(B_T)\\
\sum_{S \in \mathcal{S} : A_1 \cap S = S_1} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T)\\
\int_{B_{S_1}} p(S_1, \mathbf{x}_1) \left( \sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) \right) \mu_{S_1}(d \mathbf{x}_1) &= p(B_T) [\mbox{Fubini / Tonnelli?}]\\
\end{align}
It suffices to show that
\begin{align}
\sum_{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}} \int_{\otimes_{a \in S_2 X_a}} q_{(S_1,\mathbf{x}_1)}(S_2, \mathbf{x}_2) \mu_{S_2}(d \mathbf{x}_2) = 1 \mbox{ for all } (S_1, \mathbf{x}_1) \in T
\end{align}
It suffices to show that for each $(S_1, \mathbf{x}_1) \in T$, we have that $q_{(S_1, \mathbf{x}_1)}$ is the density of a (terminating with probability one) probabilistic program with structruers $\mathcal{S}_2$ such that
\[
\mathcal{S}_2 = \{S_2 \subseteq A_2 : S_1 \cup S_2 \in \mathcal{S}\}
\]

\[
B_1 = \cup_{S_1 \in \mathcal{S}_1} \{(S_1, \mathbf{x}_1) : \mathbf{x}_1 \in B_{S_1}\}\;\;\; \mbox{where} \;\; B_{S_1} \in \otimes_{a \in S_1} \Sigma_{a} \;\;\; [\mbox{Measurable set } B_1]
\]
% TODO the proof above should have a sum over S_1

% NOTE for each $t$ there is a different reference measure...
% we need to restrict q so that for each discrete part of $t$, the reference measure is the same
% (which implies that the same sets of random choices are made?)
% that will allow us to
